Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-07-12 AnalyticStructure_main.tex
update

% notation for standard math
\renewcommand{\d}{\partial}
\newcommand{\half}{\frac{1}{2}}
\newcommand{\dd}{{\rm d}}
\renewcommand{\r}{{\bf r}}
\newcommand{\br}{\bar{\bf r}}
\newcommand{\x}{{\bf x}}
\newcommand{\eps}{\epsilon}
\newcommand{\bomega}{\bar\omega}
\newcommand{\bbomega}{\bar{\bomega}}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\intdef}[3]{\int_{#1}^{#2} \dd {#3}}
\newcommand{\intover}[1]{\int_{-\infty}^{+\infty} \dd {#1}}
\newcommand{\sint}{\mathrlap{\displaystyle\int}
\mathrlap{\textstyle\sum}
\phantom{\mathrlap{\displaystyle
\int}\textstyle\sum}}

@@

\subsection{The analytic structure of linear susceptibility}
Each excitation ``mode'' of the system $\op E_a$ with energy $\Omega_a$, can be associated to a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$,
entirely defined in the unoccupied subspace. 
In this way we can parametrize, without loss of generality, the excitation operators of the system as:
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
For each choice of $p,a$, these objects represent, respectively, the state in which $\ket{\psi_p}$
is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\dmnot}{\op E_a}$.
We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2}
is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry 
of the excitation.
Excited states  satisfy the normalization condition
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
\ee
By applying \eqref{ExcitationOperatorsDef1} to the above parametrization
of the excitations, we obtain the following Sternheimer-like equations for the excited states:
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\bra{\chi_p^a}\left[-\Omega_a-(\hnot - \eps_p)\right] = \bra{\psi_p} \op V'[\op E_a] \op Q_0  \;.
\end{align}
By writing these equations in a basis set of virtual states we obtain the well-known Casida's eigenvalue equation\cite{CasidaBook}
associated to the excitation energy $\Omega_a$ (see appendix \ref{casida}).
The spectrum is symmetric w.r.t the inversion of the eigenvalues 
$\Omega_a \rightarrow -\Omega_a$ and the associated excitation with negative energy are described by the transpose pair $\{\chi^a_p,\phi^a_p\}$.
We concentrate our analysis on the sector of positive energies, i.e. $\Omega_a > 0$.

@@

\begin{figure}
\includegraphics[scale=0.6]{CO_statPolvsExc.eps}
\caption{\label{co_AlphaExc}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ of the $CO$ molecule, in a box of $11.5$ \AA,  w.r.t to excitations considered in the integral of Eq.~\eqref{sminus2}. The reference value obtained from Eq.~\eqref{staticalpha} is represented as a horizontal red dashed line. The value of the IP energy is indicated by a vertical black dashed line. For each excitation energy, the values of the oscillator strengths $w^a_{zz}$ of Eq.~\eqref{sminus2} are also plotted. }
\end{figure}
Hence, contrary to GS, \emph{for LR calculation,
even when the response density is localized, the contribution coming from the 
continuum sector of the excitations can never be neglected}. 
In other terms, we \emph{cannot} exclude that a fluctuation state, even if genuinely localized, will have a nonzero projection
in the essential part of the Liouvillian spectrum.
This fact shows that it is \emph{impossible} to restrict the spectral representation
of $\chi(\omega)$ only localized, discrete, \emph{code-independent} excitations, even for representing LR results in the $\omega=0$ limit.

>>>>

% notation for standard math
\renewcommand{\d}{\partial}
\newcommand{\half}{\frac{1}{2}}
\newcommand{\dd}{{\rm d}}
\renewcommand{\r}{{\bf r}}
\newcommand{\br}{\bar{\bf r}}
\renewcommand{\AA}{{Angstr\"om}}
\newcommand{\x}{{\bf x}}
\newcommand{\eps}{\epsilon}
\newcommand{\bomega}{\bar\omega}
\newcommand{\bbomega}{\bar{\bomega}}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\intdef}[3]{\int_{#1}^{#2} \dd {#3}}
\newcommand{\intover}[1]{\int_{-\infty}^{+\infty} \dd {#1}}
\newcommand{\sint}{\mathrlap{\displaystyle\int}
\mathrlap{\textstyle\sum}
\phantom{\mathrlap{\displaystyle
\int}\textstyle\sum}}

@@

\subsection{The analytic structure of linear susceptibility}
Each excitation ``mode'' of the system $\op E_a$ with energy $\Omega_a$, can be associated to a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$,
entirely defined in the unoccupied subspace. 
In this way we can parametrize, without loss of generality, the excitation operators of the system as:
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
For each choice of $\{\phi^a_p,\chi^a_p\}_{p \in a}$, these objects represent, respectively, the state in which $\ket{\psi_p}$
is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\dmnot}{\op E_a}$.
We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2}
is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry 
of the excitation.
Excited states  satisfy the normalization condition
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
\ee
By applying \eqref{ExcitationOperatorsDef1} to the above parametrization
of the excitations, we obtain the following Sternheimer-like equations for the excited states:
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\bra{\chi_p^a}\left[-\Omega_a-(\hnot - \eps_p)\right] = \bra{\psi_p} \op V'[\op E_a] \op Q_0  \;.
\end{align}
By writing these equations in a basis set of virtual states we obtain the well-known Casida's eigenvalue equation\cite{CasidaBook}
associated to the excitation energy $\Omega_a$ (see appendix \ref{casida}).
The spectrum is symmetric w.r.t the inversion of the eigenvalues 
$\Omega_a \rightarrow -\Omega_a$ and the associated excitation with negative energy are described by the transpose pair $\{\chi^a_p,\phi^a_p\}$.
We concentrate our analysis on the sector of positive energies, i.e. $\Omega_a > 0$.

@@

\begin{figure}
\includegraphics[scale=0.6]{CO_statPolvsExc.eps}
\caption{\label{co_AlphaExc}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ of the $CO$ molecule, in a box of $11.5$ \AA,  
w.r.t the excitations considered in the integral of Eq.~\eqref{sminus2}. The reference value obtained from Eq.~\eqref{staticalpha} is represented as a 
horizontal red dashed line. The value of the IP energy is indicated by a vertical black dashed line. For each excitation energy, the values of the 
oscillator strengths $w^a_{zz}$ of Eq.~\eqref{sminus2} are also plotted. }
\end{figure}
Hence, contrary to GS, \emph{for LR calculation,
even when the response density is localized, the contribution coming from the 
continuum sector of the excitations can never be neglected}. 
In other terms, we \emph{cannot} exclude that a fluctuation state, even if genuinely localized, will have a nonzero projection
in the essential part of the Liouvillian spectrum.
This fact shows that it is \emph{impossible} to restrict the spectral representation
of $\chi(\omega)$ only localized, discrete, \emph{code-independent} excitations, even for representing LR results in the $\omega=0$ limit.
